Gender, Age, and Peer Influence on Food Incentives in Elementary Schools

### 2024-02-20 outputs/paper/paper.qmd
Finished adding in the text for the intro,data,results sections

---
title: "Gender, Age, and Peer Influence on Food Incentives in Schools"
author: 
  - Samantha Barfoot
  - Brooklin Becker
  - Mehrnoush Mohammadi
thanks: "Code and data are available at: https://github.com/SamanthaBarfoot/food-incentive-analysis.git The Social Science Reproduction DOI is: https://doi.org/10.48152/ssrp-s4sw-c494"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

In this paper, we will provide a deeper analysis and reproduction of the paper, *Incentives and Unintended Consequences: Spillover Effects in Food Choice* [@angelucci_incentives_2019]. Replications of Table 1, Figure 2 and 3 were also completed from the raw data in which the original paper is based. The original study was conducted on 1631 students from various elementary schools in a low-income neighborhood in Chicago, IL, from kindergarten to grade 8. The purpose of the study was to determine whether there were spillover effects of students making decisions based on their peers' decisions as well as the incentives given to their peers. The study consisted of two stages. In the first stage (G1), each student was given a card from a deck which offered the choice of cookies or grapes, where the grapes sometimes came with a prize as an incentive to be chosen. In the second stage (G2), each student was given the opportunity to change their choice of food based on the initial decisions of their peers. Students were sat in tables of sizes up to 10 students, in which the proportion of cards in the deck that contained incentivized grapes were 0%, 50%, or 100%. Furthermore, some tables were designated as public treatment, in which students were able to see the incentive status of their peers, and some tables were designated as private treatment, in which students were unable to see the incentive status of their peers.

The remainder of this paper is structured as follows. In @sec-data, we compile the subset of data that strictly includes the students that took part in the public treatment study and not the private treatment study. Students were offered more information to advise their final decisions by being able to see the incentive status of their peers. In @sec-results, we dive deeper into analyzing possible factors that may influence a student's final food choice in stage G2. We leverage `ggplot2` by creating bar graphs to represent the grouped data for the various factors that were analyzed. Finally, in @sec-discussion, we summarize our results and also identify other factors or biases that may have affected the results, on a qualitative or quantitative level.

@@

The original study data included 1631 students, in which the incentive status, public or private, table size, table incentive proportion, sex, race, eligibility for free lunches, and food choices for both stages G1 and G2 was compiled for each student. By filtering out the original data to only show students who participated in the public treatment, the sample size reduces from 1631 students to 883 students, representing roughly 54% of the original sample size. @tbl-table1 summarizes all the students at public treatment tables, with a subset of columns displayed representing some of the factors to be analyzed in @sec-results. Comments regarding data cleaning methods are mentioned in @sec-discussion.

@@

In @tbl-table1, the results are generally consistent across the 4 subgroups, in which there is low variance among the values in the various columns. This more-or-less even distribution of the races, genders and grades of students across tables of varying incentive levels speaks to the efficacy of the randomness that was intended in the original study. In the next section, we will dive into analyzing the correlation of several individual factors and the choices made in stage G2.

@@

To determine whether there is a certain characteristic in students that might influence the decision made in stage G2, particularly their grade, gender, or their eligibility for free lunches, bar graphs were created that group the public student’s data based on each of the chosen factors.

To begin our analysis, we created a graph, @fig-foodChoices, of the frequency of students that either switched their final food choice or kept the same food choice, grouped by the grade of the student. Grade 0 represents students in kindergarten. @fig-foodChoices gives an overview of the public data across all levels of table incentivization, in which students were able to make their final choice based on their peers' initial choices and incentive statuses.
\newpage

@@

It is evident that when aggregating all the public tables, including the incentivized and non-incentivized students, that the majority of students across each grade kept their initial food choice as their final food choice. Additionally, the number of students that switched from one food to the other was approximately even for both scenarios (56 switched from Cookie to Grape, 66 switched from Grape to Cookie).

@@

We can see in @fig-lunchStatus that the proportion of students choosing the incentivized grapes that are eligible for free lunches is 94%. We can compare this to the expected value of 86% since the proportion of all students in public tables whom are eligible for free lunches is 0.86 as shown in @tbl-table1.

We can also examine the proportion of incentivized students that chose grapes compared to cookies, grouped by grade in @fig-grade.

@@

We observe that from kindergarten to grade 8, the proportion of students choosing cookies in the final stage is around 60-75%, with the majority of grades being closer to 75%. However, for students in grade 8, the proportion of students choosing cookies was less than the proportion of students choosing the incentivized grapes, as only roughly 40% of students chose the cookies. This is a considerable drop compared to all of the lower grades. This may indicate that the older students are more inclined to choose the food option which also gives them the opportunity to win a prize, and that they are less influenced by the food option itself.

Finally, @fig-gender compares the food choices of boys and girls students who were incentivized to choose between grapes and cookies. It appears that there is not much difference between boys and girls when it comes to their choices between grapes and cookies. Both genders have a slight preference for grapes and the percentage of students who choose grapes or cookies is almost the same for both genders. From our @fig-gender analysis, summary statistics were calculated, which indicate that, on average, around 31% of boys and 35% of girls chose grapes as their final food option, while approximately 69% of boys and 65% of girls chose cookies. These statistics highlight the balanced distribution of food choices among genders within the incentivized students, suggesting that there is limited gender-based influence on decision-making in this context. In conclusion, it appears that isolating based on gender has the least variance or impact on the final food choice among incentivized students compared to the other factors analyzed.

@@

We will now also create two grouped bar graphs which show the progression of the students' food choices from stage 1 to stage 2; first for public, non-incentivized students, then followed by a graph for public, incentivized students.

@@

@@

We view from these two graphs that the impact of incentivization leads to a higher proportion of students who choose the grapes in the final stage compared to the cookies. We will discuss this further in the discussion section.

>>>>

---
title: "Gender, Age, and Peer Influence on Food Incentives in Elementary Schools"
author: 
  - Samantha Barfoot
  - Brooklin Becker
  - Mehrnoush Mohammadi
thanks: "Code and data are available at: https://github.com/SamanthaBarfoot/food-incentive-analysis.git The Social Science Reproduction DOI is: https://doi.org/10.48152/ssrp-s4sw-c494"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

In this paper, we will provide a deeper analysis and reproduction of the paper, *Incentives and Unintended Consequences: Spillover Effects in Food Choice* (add reference). Replications of Table 1, Figures 2 and 3 were also completed from the raw data on which the original paper is based. The original study was conducted on 1631 students from various elementary schools in a low-income neighborhood in Chicago, IL, from kindergarten to grade 8. The purpose of the study was to determine whether there were spillover effects of students making decisions based on their peers' decisions as well as the incentives given to their peers. The study consisted of two stages. In the first stage (G1), each student was given a card from a deck which offered the choice of cookies or grapes, where the grapes sometimes came with a prize as an incentive to be chosen. In the second stage (G2), each student was given the opportunity to change their choice of food based on the initial decisions of their peers. Students were sat in tables of sizes up to 10 students, in which the proportion of cards in the deck that contained incentivized grapes were 0%, 50%, or 100%. Furthermore, some tables were designated as public treatment, in which students were able to see the incentive status of their peers, and some tables were designated as private treatment, in which students were unable to see the incentive status of their peers.

The estimand of our paper is the correlation between age group/gender and students’ food choices. Specifically, we examine how the choices are affected when students are provided with incentives and after viewing their peers choices.

The remainder of this paper is structured as follows. In @sec-data, we compile the subset of data that strictly includes the students who took part in the public treatment study and not the private treatment study. Students were offered more information to advise their final decisions by being able to see the incentive status of their peers. In @sec-results, we dive deeper into analyzing possible factors that may influence a student's final food choice in stage G2. We leverage `ggplot2` by creating bar graphs to represent the grouped data for the various factors that were analyzed. Finally, in @sec-discussion, we summarize our results and also identify other factors or biases that may have affected the results on a qualitative or quantitative level.

@@

The original study data included 1631 students, in which the incentive status, public or private, table size, table incentive proportion, sex, race, eligibility for free lunches, and food choices for both stages G1 and G2 were compiled for each student. Most of our analysis focused on students participating in the public treatment. By filtering out the original data to only show students who participated in the public treatment, the sample size was reduced from 1631 students to 883 students, representing roughly 54% of the original sample size.

The goal of the original study was to measure the effect of incentives on the students’ food choices. To do this @angelucci_incentives_2019 collected their data via a field study of elementary school students in a low-income neighbourhood in Chicago who they observed while eating lunch in school cafeterias. The study was conducted in a controlled environment and allowed the participants to choose where they sat to simulate an environment that was as natural as possible. The study was conducted across nine different schools and was administered consistently across each school. Students were allowed to sit at tables of 3-10 people. Data was collected after the students collected their food and sat at their tables, after which the researchers joined their tables and read from a script, asking questions and taking notes. Researchers would additionally record the students’ choices. Students were also assigned ID numbers to better track their food choices and aid in data collection. At the end of the study, the researchers also conducted surveys amongst the students with further questions to gain a better understanding of the social dynamics within the school.

The raw dataset analyzed in this paper contained 1631 entries with 31 different variables, where each row represented one student. The data provided in the replication package was very organized with few errors or unrealistic values. As a result, very little filtering of the rows was needed. However, we did do some cleaning to create cleaned data sets that contained only the specific data we needed to create our figures and tables. Methods such as `filter()`, `mutate()`, and `select()` were used to perform data cleaning.

Our analysis did not use all of the variables from the raw data file. Specifically, we used *incentive* which represents if the student was provided an incentive, *private* which represents if the student was part of the public or private treatment, their *grade*, the *table size* and *male* which indicates their gender. We additionally used *grape* and *grape2* which represent the students initial and final food choice. We also used the variables *Black* and *Hispanic* which represents the students race and ethnicity. Lastly, we used *free* which indicates if the student is a part of the free lunch program.

We created two variables called *Switched* and *Stayed* for use in @fig-foodChoices, @fig-publicNoIncentive, and @fig-publicIncentive. *Switched* is used to determine which students switched from grapes to cookies, cookies to grapes, or did not change their food choice at all. This variable was created from *grape* and *grape2*. The variable *Stayed* was also created from *grape* and *grape2* and was used to determine which students stayed with their initial food choices.

We created @tbl-table1 to summarize all the students at public treatment tables, with a subset of columns displayed representing some of the factors to be analyzed in @sec-results.

@@

In @tbl-table1, the results are generally consistent across the four subgroups, in which there is low variance among the values in the various columns. This more-or-less even distribution of the races, genders and grades of students across tables of varying incentive levels speaks to the efficacy of the randomness that was intended in the original study. We will now dive into analyzing the correlation of several individual factors and the choices made in stage G2.

@@

Bar graphs were created to determine whether there are certain characteristics in students that might influence their decision made in stage G2. These characteristics include their grade, gender, or eligibility for free lunches. The public student data was grouped based on each of these factors.

To begin our analysis, we created @fig-foodChoices, which represents the frequency of students that either switched their final food choice or kept the same food choice, grouped by the grade of the student. Grade 0 represents students in kindergarten. @fig-foodChoices gives an overview of the public data across all levels of table incentivizes, in which students were able to make their final choice based on their peers' initial choices and incentive statuses.

@@

It is evident that when aggregating all the public tables, including the incentivized and non-incentivized students, the majority of students across each grade kept their initial food choice as their final food choice. Additionally, the number of students that switched from one food to the other was approximately equal for both scenarios (56 switched from cookie to grape, 66 switched from grape to cookie).

@@

We can see in @fig-lunchStatus that the proportion of students choosing the incentivized grapes that are eligible for free lunches is roughly 94% (15/16). We can compare this to the expected value of 87% since the proportion of all students in public tables who are eligible for free lunches is 0.87, as shown in @tbl-table1.

We can also examine the proportion of incentivized students that chose grapes compared to cookies, grouped by grade in @fig-grade. We observe that from kindergarten to grade 8, the proportion of students choosing cookies in the final stage is around 60-75%, with the majority of grades being closer to 75%. However, for students in grade 8, the proportion of students choosing cookies was less than the proportion of students choosing the incentivized grapes, as only roughly 40% of students chose the cookies. This is a considerable drop compared to all of the lower grades. This may indicate that the older students are more inclined to choose the food option which also gives them the opportunity to win a prize, and that they are less influenced by the food option itself.

@@

@fig-gender compares the food choices of boys and girls students who were incentivized to choose between grapes and cookies. It appears that there is not much difference between boys and girls when it comes to their choices between grapes and cookies. Both genders have a slight preference for grapes and the percentage of students who choose grapes or cookies is almost the same for both genders. On average, around 31% of boys and 35% of girls chose grapes as their final food option, while approximately 69% of boys and 65% of girls chose cookies. This highlights the balanced distribution of food choices among genders within the incentivized students, suggesting that there is limited gender-based influence on decision-making in this context. In conclusion, it appears that isolating based on gender has the least variance or impact on the final food choice among incentivized students compared to the other factors analyzed.

@@

Lastly, we compared the proportion of students in the public treatment who received no incentive in three groups: those who stayed with their initial choice of cookies, those who stayed with their initial choice of grapes and those who changed their choice. This can be seen in @fig-publicNoIncentive. We found that students stayed with their choice of cookie more than the other two options. Overall, most students stayed with their initial choice if they were in the public treatment and received no incentive.

@@

We created a similar graph, as seen in @fig-publicIncentive, but with students who received the public treatment and received an incentive. In this comparison we found that students mostly stayed with their choice of grape. Therefore, when provided with an incentive, students appear to pick grapes and stay with that choice. Whereas, with no incentive, they pick the cookie and stay with that choice.

@@

We view from these two graphs that the impact of incentivization leads to a higher proportion of students who choose the grapes in the final stage compared to the cookies.
